---
title: "Catalog Backends: Choosing Your Metadata Store"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Catalog Backends: Choosing Your Metadata Store}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette provides an in-depth look at DuckLake catalog backends - the database that stores metadata about your tables, schemas, snapshots, and file references. Choosing the right backend is crucial for performance, concurrency, and operational simplicity.

---

## What is the Catalog?

DuckLake separates **data** (Parquet files) from **metadata** (table definitions, snapshots, file tracking). The metadata lives in a **catalog database**.

```
┌─────────────────────────────────────────────────────────┐
│                    DuckLake Architecture                 │
├─────────────────────────────────────────────────────────┤
│                                                          │
│  ┌──────────────────┐      ┌──────────────────────────┐ │
│  │  Catalog (Meta)  │      │    Data (Parquet)        │ │
│  │                  │      │                          │ │
│  │  • Table schemas │      │  📁 schema/              │ │
│  │  • Snapshots     │◄────►│    📁 table/            │ │
│  │  • File refs     │      │      📄 data-001.parquet│ │
│  │  • Partitions    │      │      📄 data-002.parquet│ │
│  │  • Lineage       │      │                          │ │
│  └──────────────────┘      └──────────────────────────┘ │
│         ▲                                                │
│         │                                                │
│    Which database?                                       │
│    • DuckDB                                              │
│    • SQLite                                              │
│    • PostgreSQL                                          │
│    • Quack (EXPERIMENTAL)                                │
│    • (Future: SQL Server)                                │
│                                                          │
└─────────────────────────────────────────────────────────┘
```

The catalog database choice affects:

- **Concurrency** - How many users can read/write simultaneously
- **Infrastructure** - Server required vs. just a file
- **Performance** - Metadata query speed
- **Operational complexity** - Backup, monitoring, maintenance

---

## Current Backends

### DuckDB (Native)

```{r}
db_connect(
 catalog_type = "duckdb",
  metadata_path = "catalog.ducklake",
  data_path = "/data/lake"
)
```

| Aspect | Details |
|--------|---------|
| **Storage** | Single `.ducklake` file |
| **Concurrency** | Single client only |
| **Infrastructure** | None - just a file |
| **Best for** | Personal use, development, testing |

**Advantages:**

- Simplest setup - just a file path
- Fastest metadata operations (native DuckDB)
- No external dependencies
- Perfect for local development and prototyping

**Disadvantages:**

- **Single client limitation** - only one connection at a time
- File locking prevents concurrent access
- Not suitable for shared environments

**When to use:**

- Local development and testing
- Single-user analytical workflows
- CI/CD pipelines (isolated runs)
- Prototyping before moving to production

---

### SQLite

```{r}
db_connect(
  catalog_type = "sqlite",
  metadata_path = "//network/share/catalog.sqlite",
  data_path = "//network/share/data"
)
```

| Aspect | Details |
|--------|---------|
| **Storage** | Single `.sqlite` file |
| **Concurrency** | Multiple readers, single writer (with retry) |
| **Infrastructure** | None - just a file |
| **Best for** | Shared network drives, small teams |

**Advantages:**

- Still just a file - no server needed
- Multiple users can read simultaneously
- Write conflicts handled with automatic retry
- Works with existing network share permissions
- Familiar to IT teams (standard SQLite)

**Disadvantages:**

- Single writer at a time (queued with retry)
- Performance degrades over very slow network connections
- No true concurrent writes

**When to use:**

- **Recommended default for shared environments**
- Network drive deployments
- Teams where writes are infrequent
- Environments where PostgreSQL isn't available

**Concurrency model:**

```
User A reads  ───────────────────────────► OK (instant)
User B reads  ───────────────────────────► OK (instant)
User C writes ─────────────────┐
                               ├──────────► OK (gets lock)
User D writes ─────────────────┘
              waits... retries... ────────► OK (after C finishes)
```

---

### PostgreSQL

```{r}
db_connect(
  catalog_type = "postgres",
  metadata_path = "host=db.example.com dbname=ducklake user=app",
  data_path = "/data/lake"
)
```

| Aspect | Details |
|--------|---------|
| **Storage** | PostgreSQL database |
| **Concurrency** | Full multi-reader, multi-writer |
| **Infrastructure** | PostgreSQL server required |
| **Best for** | Large teams, high concurrency, enterprise |

**Advantages:**

- True concurrent read and write access
- Row-level locking - writes don't block each other
- Robust transaction support
- Enterprise features: replication, backup, monitoring
- Works with managed services (AWS RDS, Azure, GCP Cloud SQL)

**Disadvantages:**

- Requires PostgreSQL server infrastructure
- Additional operational overhead (backups, updates, monitoring)
- Network latency for metadata operations
- More complex setup

**When to use:**

- Multiple users writing frequently
- Enterprise environments with existing PostgreSQL
- When you need high availability
- Remote/distributed teams

**Connection string formats:**

```{r}
# Key-value format
db_connect(
  catalog_type = "postgres",
  metadata_path = "host=localhost port=5432 dbname=ducklake user=myuser password=secret"
)

# URI format
db_connect(
  catalog_type = "postgres",
  metadata_path = "postgresql://myuser:secret@localhost:5432/ducklake"
)
```

---

## Comparison Matrix

| Feature | DuckDB | SQLite | PostgreSQL | Quack |
|---------|--------|--------|------------|-------|
| **Setup complexity** | Trivial | Simple | Moderate | Moderate |
| **Infrastructure** | None | None | Server required | DuckDB server |
| **Concurrent readers** | 1 | Unlimited | Unlimited | Unlimited |
| **Concurrent writers** | 1 | 1 (queued) | Unlimited | Unlimited |
| **Network drive friendly** | No | Yes | N/A | N/A |
| **Cloud managed option** | No | No | Yes | No |
| **Metadata query speed** | Fastest | Fast | Network dependent | Fast |
| **Backup strategy** | Copy file | Copy file | pg_dump / replication | DuckDB backup |
| **Best team size** | 1 | 2-20 | 20+ | 2-50 |
| **Status** | Stable | Stable | Stable | EXPERIMENTAL |

---

## Security & Access Control

DuckLake's security model has two distinct layers:

1. **Catalog security** - Who can read/modify table definitions and metadata
2. **Data security** - Who can read/write the actual Parquet files

### Current Model: File System Permissions

Currently, DuckLake relies on **file system permissions** for data access control:

```
┌─────────────────────────────────────────────────────────────┐
│                   Security Architecture                      │
├─────────────────────────────────────────────────────────────┤
│                                                              │
│  Catalog (Metadata)              Data (Parquet Files)        │
│  ┌──────────────────┐           ┌──────────────────────┐    │
│  │ catalog.sqlite   │           │ /data/trade/         │    │
│  │                  │           │   imports/*.parquet  │    │
│  │ - Table schemas  │           │   exports/*.parquet  │    │
│  │ - Snapshots      │           │                      │    │
│  │ - File refs      │           │ /data/labour/        │    │
│  └────────┬─────────┘           │   employment/*.parquet│   │
│           │                     └──────────┬───────────┘    │
│           ▼                                ▼                │
│    File/DB permissions            Folder ACLs (Windows/NFS) │
│                                                              │
└─────────────────────────────────────────────────────────────┘
```

**Important limitation:** If a user can read the Parquet files directly (bypassing DuckLake), they have full access to that data. DuckLake does not encrypt data at rest or enforce row/column-level security.

### Security by Backend

| Backend | Catalog Security | Data Security |
|---------|------------------|---------------|
| **DuckDB** | File permissions only | File/folder ACLs |
| **SQLite** | File permissions only | File/folder ACLs |
| **PostgreSQL** | Full RBAC, SSL, row-level security | File/folder ACLs |

### PostgreSQL: Enhanced Security Options

PostgreSQL offers significantly more security features for the **catalog**:

```{r}
# SSL/TLS encrypted connection
db_connect(
  catalog_type = "postgres",
  metadata_path = "host=db.example.com dbname=ducklake user=app sslmode=require"
)
```

**PostgreSQL security features:**

| Feature | Description |
|---------|-------------|
| **Role-based access (RBAC)** | Grant/revoke permissions per user/role |
| **SSL/TLS encryption** | Encrypt metadata in transit |
| **Row-level security (RLS)** | Restrict which snapshots/tables users can see |
| **Audit logging** | Track who accessed/modified metadata |
| **LDAP/Kerberos/SCRAM** | Enterprise authentication integration |
| **Connection pooling** | PgBouncer for secure connection management |

**Example: Restricting schema access in PostgreSQL:**

```sql
-- Create roles for different teams
CREATE ROLE trade_team;
CREATE ROLE labour_team;

-- Grant access to specific schema metadata only
GRANT SELECT ON ducklake_schema TO trade_team
  WHERE schema_name = 'trade';

-- Users inherit permissions from their role
GRANT trade_team TO alice, bob;
```

### Future: Native DuckLake Security

DuckLake may introduce native security features in future releases:

**Potential features being discussed:**

- **Catalog-level RBAC** - Define permissions in DuckLake metadata itself
- **Schema-level grants** - `GRANT SELECT ON SCHEMA trade TO role_x`
- **Table-level grants** - `GRANT INSERT ON table TO role_y`
- **Encryption at rest** - Encrypted Parquet files with key management
- **Column masking** - Hide sensitive columns from certain users

**Current workaround for sensitive data:**

```{r}
# Store sensitive data in a separate schema with restricted folder ACLs
db_write(sensitive_data, schema = "restricted", table = "pii")

# Folder structure:
# /data/restricted/  <- Only authorized users have ACL access
#   pii/*.parquet
```

### Quack Security Considerations

When Quack becomes available as a catalog backend, it will introduce new security considerations:

```
┌─────────────────────────────────────────────────────────────┐
│                    Quack Security Model                      │
├─────────────────────────────────────────────────────────────┤
│                                                              │
│  Client ──── HTTPS ────► DuckDB Server ────► Data Files     │
│                              │                               │
│                              ▼                               │
│                    Authentication                            │
│                    Authorization                             │
│                    Audit logging                             │
│                                                              │
└─────────────────────────────────────────────────────────────┘
```

**Expected Quack security features:**

| Feature | Status |
|---------|--------|
| **HTTPS transport** | Available now |
| **Token authentication** | Available now |
| **Reverse proxy support** | Available now (nginx, Caddy) |
| **Native RBAC** | Planned |
| **Query auditing** | Planned |

**Securing Quack with a reverse proxy:**

```
# nginx configuration for Quack
server {
    listen 443 ssl;
    server_name duckdb.example.com;

    ssl_certificate /etc/ssl/certs/duckdb.crt;
    ssl_certificate_key /etc/ssl/private/duckdb.key;

    location / {
        proxy_pass http://localhost:4242;
        proxy_set_header Authorization $http_authorization;
    }
}
```

### Security Recommendations by Use Case

| Use Case | Recommended Backend | Security Approach |
|----------|---------------------|-------------------|
| **Personal/Dev** | DuckDB | Local file permissions |
| **Team, low sensitivity** | SQLite | Folder ACLs per schema |
| **Team, sensitive data** | PostgreSQL | RBAC + folder ACLs + SSL |
| **Enterprise, compliance** | PostgreSQL | Full RBAC + RLS + audit + encryption |
| **Multi-tenant** | PostgreSQL + Quack | Row-level security + API gateway |

### Data Cannot Be Fully Controlled via Catalog

**Critical understanding:** The catalog backend only controls **metadata access**. The actual data files (Parquet) are controlled by file system permissions.

```{r}
# Even if PostgreSQL restricts a user from seeing table metadata...
# ...they could still read the parquet files directly if they have folder access:

# Direct file access bypasses DuckLake entirely
arrow::read_parquet("/data/trade/imports/file-001.parquet")
```

**Mitigations:**

1. **Folder ACLs** - Ensure Parquet folders have appropriate permissions
2. **Network isolation** - Store data on restricted network shares
3. **Encryption** - Use encrypted storage (BitLocker, LUKS, cloud KMS)
4. **Monitoring** - Audit file access at the OS level

---

## Future: SQL Server Support

### Current Status

SQL Server is **not yet supported** as a DuckLake catalog backend. This is tracked in:

- [DuckLake Discussion #892](https://github.com/duckdb/ducklake/discussions/892)

### Why SQL Server?

Many enterprise environments standardise on SQL Server:

- Existing infrastructure and expertise
- Azure SQL Database / Managed Instance availability
- Corporate licensing agreements
- Integration with Microsoft ecosystem (Power BI, Azure Data Factory)

### Technical Path

The [hugr-lab/mssql-extension](https://github.com/hugr-lab/mssql-extension) provides native SQL Server connectivity for DuckDB:

- Native TDS protocol (no ODBC/JDBC required)
- Full DML support (INSERT, UPDATE, DELETE)
- Transaction support
- High-performance bulk loading via BCP protocol
- Azure AD authentication

**What's missing:** DuckLake needs to add `ducklake:mssql:` DSN support and generate SQL Server-compatible DDL.

### What SQL Server Would Offer

```{r}
# Future (once DuckLake supports it)
db_connect(
  catalog_type = "mssql",
  metadata_path = "Server=sql.example.com;Database=ducklake;User Id=app;Password=secret",
  data_path = "/data/lake"
)
```

| Feature | Benefit |
|---------|---------|
| **Enterprise integration** | Use existing SQL Server infrastructure |
| **Azure native** | Azure SQL Database, Managed Instance |
| **Familiar tooling** | SSMS, Azure Data Studio |
| **Existing permissions** | AD/Entra ID authentication |
| **High availability** | Always On, geo-replication |

### Current Workarounds

Until SQL Server support arrives:

1. **PostgreSQL** - Use Azure Database for PostgreSQL in Azure environments
2. **SQLite** - Store catalog file on Azure Files or shared storage
3. **Hybrid approach** - Keep DuckLake metadata in PostgreSQL, source data in SQL Server

---

## Quack Remote Protocol (EXPERIMENTAL)
### What is Quack?

[Quack](https://duckdb.org/2026/05/12/quack-remote-protocol) is DuckDB's client-server protocol, available since DuckDB v1.5.2:

- **RPC protocol** for DuckDB-to-DuckDB communication
- **HTTP-based** for network compatibility (port 9494 by default)
- **Native serialisation** of DuckDB's internal data vectors
- **Multi-writer support** without locking

As of DuckDB 1.5.3, DuckLake supports Quack as a catalog backend. datapond includes experimental support for this.

### How Quack Changes Everything

```
Before Quack (DuckDB catalog):                After Quack:
┌─────────────────────────────┐             ┌─────────────────────────────┐
│     Single Client Only      │             │   Multiple Clients          │
│                             │             │                             │
│  Client ──► DuckDB File     │             │  Client A ──┐               │
│             (locked)        │             │  Client B ──┼──► DuckDB     │
│                             │             │  Client C ──┘    Server     │
└─────────────────────────────┘             └─────────────────────────────┘
```

### Using Quack with datapond

datapond supports Quack as an experimental catalog backend:

```{r}
# Connect to a DuckDB server running Quack
db_connect(
  catalog_type = "quack",
  metadata_path = "quack:duckdb-server.example.com:9494/catalog.ducklake",
  data_path = "/data/lake"
)
```
You will see a warning about the experimental status. This is expected.

**Benefits:**

| Feature | Impact |
|---------|--------|
| **Multi-writer DuckDB** | Native DuckDB speed with concurrent access |
| **Inlining performance** | Much faster than SQLite/PostgreSQL for streaming |
| **Simplified stack** | DuckDB for everything - no PostgreSQL needed |
| **Edge deployment** | Run DuckDB server close to data |

### Running a Quack Server

To use Quack, you need a DuckDB server:

```bash
# Start DuckDB with Quack server enabled
duckdb -cmd "INSTALL quack; LOAD quack; CALL quack_start_server();"
```

The server listens on port 9494 by default. See the [Quack documentation](https://duckdb.org/docs/current/quack/setup/overview) for configuration options.

### Status and Timeline

| Version | Status |
|---------|--------|
| **DuckDB 1.5.2** | Quack beta available |
| **DuckDB 1.5.3** | DuckLake + Quack integration |
| **DuckDB 2.0 (Fall 2026)** | Production-ready Quack |

**Warning:** The protocol, function names, and defaults may change before v2.0. Use in production at your own risk.

### Quack Authentication

datapond supports basic token authentication for Quack:

```{r}
# Option 1: Explicit token parameter
db_connect(
  catalog_type = "quack",
  metadata_path = "quack:server:9494/catalog",
  data_path = "/data/lake",
  quack_token = "my-secret-token"
)

# Option 2: Environment variable (recommended for scripts)
Sys.setenv(QUACK_TOKEN = "my-secret-token")
db_connect(
  catalog_type = "quack",
  metadata_path = "quack:server:9494/catalog",
  data_path = "/data/lake"
)
```

### Quack Security: Future Possibilities

Quack is designed with a "bring your own security" philosophy. While datapond currently
supports only token authentication, Quack itself supports custom authentication and
authorization callbacks. Future datapond releases may support:
| Feature | Description | Status |
|---------|-------------|--------|
| **Token auth** | Simple shared secret | Implemented |
| **Environment variable** | `QUACK_TOKEN` fallback | Implemented |
| **Custom headers** | JWT/Bearer tokens via reverse proxy | Future |
| **Credential callback** | Dynamic token retrieval (e.g., from keyring) | Future |
| **LDAP/AD integration** | Enterprise directory auth | Requires server-side setup |

**Production deployment pattern:**

For production, the recommended approach is to put Quack behind a reverse proxy
(nginx, Caddy) that handles TLS termination and authentication:

```
Client ──► HTTPS ──► nginx ──► HTTP ──► Quack Server
                      │
                TLS + Auth headers
                Rate limiting
                Audit logging
```

This keeps Quack simple while leveraging battle-tested HTTP infrastructure for security.

See [Securing Quack with a Reverse Proxy](https://duckdb.org/docs/current/quack/setup/reverse_proxy)
for detailed setup instructions.

### Setting up a Quack Server

Unlike SQLite or DuckDB file-based catalogs, Quack requires a running DuckDB server process.
This section covers how to set up and run a Quack server for use with datapond.

#### Basic Server Setup

**1. Start a simple server (development/testing):**

```bash
# Start DuckDB with Quack server
duckdb -cmd "
  INSTALL quack;
  LOAD quack;
  INSTALL ducklake;
  LOAD ducklake;

  -- Create and attach the DuckLake catalog
  ATTACH 'ducklake:catalog.ducklake' AS lake (DATA_PATH '/data/lake');

  -- Start the Quack server
  SELECT * FROM quack_serve();
"
```
This prints the connection details:

```
┌─────────────────────────────────┬────────────────────────────────┬──────────────────────────┐
│ listen_uri                      │ http_url                       │ token                    │
├─────────────────────────────────┼────────────────────────────────┼──────────────────────────┤
│ quack:localhost:9494            │ http://localhost:9494          │ abc123xyz...             │
└─────────────────────────────────┴────────────────────────────────┴──────────────────────────┘
```

**2. Set a custom token:**

```bash
duckdb -cmd "
  INSTALL quack; LOAD quack;
  INSTALL ducklake; LOAD ducklake;

  -- Set a known token (min 4 characters)
  SET quack_token = 'my-secret-token';

  ATTACH 'ducklake:catalog.ducklake' AS lake (DATA_PATH '/data/lake');
  SELECT * FROM quack_serve();
"
```

**3. Connect from R:**

```{r}
db_connect(
  catalog_type = "quack",
  metadata_path = "quack:localhost:9494/lake",
  data_path = "/data/lake",
  quack_token = "my-secret-token"
)
```

#### Running as a Service (Linux)

For production, run the Quack server as a systemd service:

**1. Create a startup script** (`/opt/duckdb/start-quack.sql`):

```sql
INSTALL quack;
LOAD quack;
INSTALL ducklake;
LOAD ducklake;

-- Set authentication token from environment
SET quack_token = getenv('QUACK_TOKEN');

-- Attach the DuckLake catalog
ATTACH 'ducklake:/opt/duckdb/catalog.ducklake' AS lake
  (DATA_PATH '/data/lake');

-- Start server (blocks)
SELECT * FROM quack_serve(port := 9494);
```

**2. Create systemd service** (`/etc/systemd/system/duckdb-quack.service`):

```ini
[Unit]
Description=DuckDB Quack Server
After=network.target

[Service]
Type=simple
User=duckdb
Group=duckdb
Environment=QUACK_TOKEN=your-secret-token-here
WorkingDirectory=/opt/duckdb
ExecStart=/usr/local/bin/duckdb -init /opt/duckdb/start-quack.sql
Restart=always
RestartSec=10

[Install]
WantedBy=multi-user.target
```

**3. Enable and start:**

```bash
sudo systemctl daemon-reload
sudo systemctl enable duckdb-quack
sudo systemctl start duckdb-quack

# Check status
sudo systemctl status duckdb-quack
```

#### Running with Docker

```dockerfile
# Dockerfile
FROM ubuntu:22.04

RUN apt-get update && apt-get install -y wget unzip
RUN wget https://github.com/duckdb/duckdb/releases/download/v1.5.3/duckdb_cli-linux-amd64.zip \
    && unzip duckdb_cli-linux-amd64.zip -d /usr/local/bin/

COPY start-quack.sql /opt/duckdb/
VOLUME ["/data"]
EXPOSE 9494

CMD ["duckdb", "-init", "/opt/duckdb/start-quack.sql"]
```

```bash
# Build and run
docker build -t duckdb-quack .
docker run -d \
  -p 9494:9494 \
  -v /path/to/data:/data \
  -e QUACK_TOKEN=my-secret-token \
  --name quack-server \
  duckdb-quack
```

#### Production Setup with nginx

For production, put nginx in front of Quack for TLS and additional security:

**1. nginx configuration** (`/etc/nginx/sites-available/quack`):

```nginx
server {
    listen 443 ssl http2;
    server_name duckdb.example.com;

    ssl_certificate /etc/letsencrypt/live/duckdb.example.com/fullchain.pem;
    ssl_certificate_key /etc/letsencrypt/live/duckdb.example.com/privkey.pem;

    location / {
        proxy_pass http://127.0.0.1:9494;
        proxy_http_version 1.1;
        proxy_set_header Host $host;
        proxy_set_header X-Real-IP $remote_addr;
        proxy_set_header X-Forwarded-For $proxy_add_x_forwarded_for;
        proxy_set_header X-Forwarded-Proto $scheme;

        # Pass through authentication header
        proxy_set_header Authorization $http_authorization;

        # Timeouts for long queries
        proxy_read_timeout 300s;
        proxy_connect_timeout 75s;
    }
}
```

**2. Connect via HTTPS:**

```{r}
# Quack client assumes HTTPS for non-localhost URIs
db_connect(
  catalog_type = "quack",
  metadata_path = "quack:duckdb.example.com/lake",
  data_path = "/data/lake",
  quack_token = "my-secret-token"
)
```

#### Quick Deployment (AWS)

DuckDB provides a one-click CloudFormation template for AWS:

1. Go to [Deploying Quack](https://duckdb.org/docs/current/quack/setup/deployment)
2. Click "Launch Stack" for your region
3. Configure instance size, token, and data path
4. Stack provisions EC2 + nginx + Let's Encrypt automatically

#### Monitoring and Troubleshooting

**Check server status:**

```bash
# Is the server running?
curl http://localhost:9494/health

# View connections (from DuckDB CLI on server)
SELECT * FROM quack_connections();
```

**Common issues:**

| Issue | Solution |
|-------|----------|
| Connection refused | Check server is running, firewall allows port 9494 |
| Authentication failed | Verify token matches between client and server |
| Timeout on large queries | Increase `proxy_read_timeout` in nginx |
| "Extension not found" | Run `INSTALL quack; INSTALL ducklake;` on server |

---

## Geospatial Data Considerations

### Native Geometry Support

DuckLake 1.0 includes native support for geometry types via the DuckDB spatial extension:

```{r}
# Load spatial extension
DBI::dbExecute(con, "INSTALL spatial; LOAD spatial;")

# Write geospatial data
db_write(
  geo_data,
  table = "boundaries",
  mode = "overwrite"
)
```

**Supported geometry types:**

- POINT, LINESTRING, POLYGON
- MULTIPOINT, MULTILINESTRING, MULTIPOLYGON
- GEOMETRYCOLLECTION

### GeoParquet Storage

DuckDB reads and writes [GeoParquet](https://geoparquet.org/) natively:

```{r}
# Spatial data stored as GeoParquet
boundaries <- db_read(table = "boundaries")

boundaries |>
  filter(ST_Intersects(geometry, ST_Point(-6.26, 53.35))) |>
  collect()
```

### Formats That Can't Be Stored Directly

Some geospatial formats don't fit the tabular model:

| Format | Issue | Solution |
|--------|-------|----------|
| **Raster (GeoTIFF, COG)** | Grid data, not rows | Store file reference + metadata |
| **Point clouds (LAS/LAZ)** | Billions of points | Store summary + file path |
| **3D models (CityGML)** | Complex hierarchical | Store metadata + file reference |
| **Vector tiles (MVT)** | Pre-rendered tiles | Store tile index + paths |

### Metadata-Only Approach

For unsupported formats, store metadata in DuckLake while keeping files external:

```{r}
# Create a metadata table for raster files
raster_catalog <- data.frame(
  file_id = 1:3,
  file_path = c(
    "/data/rasters/dem_2024.tif",
    "/data/rasters/landcover_2024.tif",
    "/data/rasters/ndvi_2024_q1.tif"
  ),
  format = "GeoTIFF",
  crs = "EPSG:2157",
  bounds_wkt = c(
    "POLYGON((-10.5 51.4, -5.5 51.4, -5.5 55.4, -10.5 55.4, -10.5 51.4))",
    "POLYGON((-10.5 51.4, -5.5 51.4, -5.5 55.4, -10.5 55.4, -10.5 51.4))",
    "POLYGON((-10.5 51.4, -5.5 51.4, -5.5 55.4, -10.5 55.4, -10.5 51.4))"
  ),
  resolution_m = c(10, 30, 10),
  acquisition_date = as.Date(c("2024-01-15", "2024-06-01", "2024-03-15")),
  file_size_mb = c(450, 120, 380),
  description = c(
    "Digital Elevation Model - 10m resolution",
    "CORINE Land Cover 2024",
    "NDVI composite Q1 2024"
  )
)

db_write(raster_catalog, table = "raster_catalog", mode = "overwrite")

# Document the catalog
db_describe(
  table = "raster_catalog",
  description = "Metadata catalog for raster files stored externally",
  owner = "GIS Team",
  tags = c("geospatial", "raster", "metadata-only")
)
```

### Querying Spatial Metadata

```{r}
# Find rasters that intersect an area of interest
raster_catalog <- db_read(table = "raster_catalog")

aoi <- "POLYGON((-7 53, -6 53, -6 54, -7 54, -7 53))"

raster_catalog |>
  filter(ST_Intersects(ST_GeomFromText(bounds_wkt), ST_GeomFromText(aoi))) |>
  select(file_path, description, acquisition_date) |>
  collect()
```

### Hybrid Architecture for Geospatial

```
┌─────────────────────────────────────────────────────────────┐
│                  Geospatial Data Architecture               │
├─────────────────────────────────────────────────────────────┤
│                                                             │
│  DuckLake (Tabular + Vector Geometry)                       │
│  ┌─────────────────────────────────────────────────────┐   │
│  │  boundaries     (POLYGON geometries)                 │   │
│  │  points_of_interest (POINT geometries)               │   │
│  │  road_network   (LINESTRING geometries)              │   │
│  │  raster_catalog (metadata only - file references)    │   │
│  │  pointcloud_index (metadata only - file references)  │   │
│  └─────────────────────────────────────────────────────┘   │
│                           │                                 │
│                           ▼                                 │
│  External Storage (Non-tabular formats)                     │
│  ┌─────────────────────────────────────────────────────┐   │
│  │  /data/rasters/          (GeoTIFF, COG)              │   │
│  │  /data/pointclouds/      (LAS, LAZ)                  │   │
│  │  /data/tiles/            (MVT, PMTiles)              │   │
│  └─────────────────────────────────────────────────────┘   │
│                                                             │
└─────────────────────────────────────────────────────────────┘
```

---

## Decision Guide

### Quick Reference

```
Are you the only user?
├─ Yes ──► DuckDB (simplest)
└─ No
   └─ Do you need concurrent writes?
      ├─ No (occasional writes) ──► SQLite
      └─ Yes (frequent writes)
         └─ Do you have SQL Server and need it?
            ├─ Yes ──► Wait for MSSQL support, use PostgreSQL interim
            └─ No ──► PostgreSQL
```

### Migration Path

Start simple, scale up as needed:

```
Development     Team Pilot      Production       Enterprise
    │               │               │                │
    ▼               ▼               ▼                ▼
 DuckDB ────► SQLite ────► PostgreSQL ────► PostgreSQL HA
                                      └───► Quack (future)
```

The data files (Parquet) remain unchanged - only the catalog backend changes.

---

## Summary

| Backend | Best For | Key Limitation |
|---------|----------|----------------|
| **DuckDB** | Solo development | Single client |
| **SQLite** | Small teams, network drives | Single writer |
| **PostgreSQL** | Enterprise, high concurrency | Requires server |
| **Quack** | DuckDB-native multi-writer | EXPERIMENTAL (stable in v2.0) |
| **SQL Server** | Microsoft shops | Not yet supported |

For most users starting out, **SQLite is the recommended default** - it's simple, requires no infrastructure, and handles typical team usage patterns well. Once Quack reaches production status (DuckDB 2.0, Fall 2026), it may become the preferred option for teams wanting multi-writer without PostgreSQL infrastructure.

---

## References

- [DuckLake Catalog Documentation](https://ducklake.select/docs/stable/duckdb/usage/choosing_storage)
- [Quack Protocol Announcement](https://duckdb.org/2026/05/12/quack-remote-protocol)
- [DuckDB Spatial Extension](https://duckdb.org/docs/extensions/spatial)
- [GeoParquet Specification](https://geoparquet.org/)
- [SQL Server Support Discussion](https://github.com/duckdb/ducklake/discussions/892)
- [MSSQL Extension](https://github.com/hugr-lab/mssql-extension)
