---
title: "Code Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Code Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette explains how `datapond` works under the hood. It walks through each file in the package, explaining what the code does and why it's designed that way.

---

## Package Structure

```
datapond/
├── DESCRIPTION          # Package metadata
├── NAMESPACE            # Exports and imports
├── LICENSE              # MIT license
├── R/
│   ├── db_connect.R     # Connection management
│   ├── db_read.R        # Read functions
│   ├── db_write.R       # Write functions
│   ├── upsert.R         # MERGE operations
│   ├── metadata.R       # Snapshot/catalog queries
│   ├── discovery.R      # List/exists functions
│   ├── maintenance.R    # Vacuum, rollback, diff, compaction
│   ├── advanced.R       # Clustering, inlining, Iceberg export
│   ├── arrow.R          # Arrow integration (read/write)
│   ├── docs.R           # Documentation & data dictionary
│   ├── preview.R        # Write preview functions
│   ├── browser.R        # Shiny browser launcher
│   ├── browser_ui.R     # Shiny browser UI module
│   ├── browser_server.R # Shiny browser server module
│   └── zzz.R            # Package load/unload hooks
└── vignettes/
    ├── concepts.Rmd     # Conceptual background
    ├── catalog-backends.Rmd  # Catalog backend options
    └── code-walkthrough.Rmd  # This file
```

---

## Core Design Decisions

### 1. Singleton Connection Pattern

The package maintains a single connection stored in a private environment:
```{r}
.db_env <- new.env(parent = emptyenv())
```

**Why?**

- Prevents users accidentally creating multiple connections
- Simplifies the API (no need to pass connection objects around)
- Ensures cleanup happens properly

All functions retrieve the connection via `.db_get_con()` rather than creating new ones.

### 2. Input Validation

All user inputs are validated early with clear error messages:
```{r}
.db_validate_name <- function(x, arg = "name") {
  # Must be single non-empty string
  # Only allows A-Z a-z 0-9 _ -
  # Prevents SQL injection and path traversal
}
```

This catches problems like `table = "../../../etc/passwd"` before they cause harm.

### 3. Lazy Evaluation with DuckDB

Read functions return **lazy** dplyr tables, not collected data:
```{r}
dplyr::tbl(con, dplyr::sql(query))
```

**Why?**

- Large datasets aren't loaded into memory until needed
- DuckDB can optimise the full query (filters, joins, aggregations)
- Users call `collect()` when they're ready to bring data into R

### 4. DuckLake SQL Parameter Syntax

DuckLake uses two different syntaxes for named parameters depending on the function:

**`=>` for DuckLake extension functions:**
```{r}
# Functions like ducklake_merge_adjacent_files, ducklake_list_files, etc.
CALL ducklake_merge_adjacent_files('catalog', 'table', schema => 'main')
FROM ducklake_list_files('catalog', 'table', schema => 'main')
```

**`:=` for catalog-level set_option calls:**
```{r}
# Setting options like data_inlining_row_limit
CALL catalog.set_option('data_inlining_row_limit', 100, table_name := 'events')
```

This distinction matters because using the wrong syntax will cause SQL errors. The package handles this internally, but it's important to understand when reading the generated SQL.

---

## File-by-File Walkthrough

---

## `R/db_connect.R` - Connection Management

This file handles connecting to and disconnecting from DuckLake.

### Private Environment and Helpers

```{r}
.db_env <- new.env(parent = emptyenv())
```

This creates an isolated environment that persists for the R session. We store:

- `con` - the DuckDB connection object
- `data_path` - root path for parquet files
- `catalog` - DuckLake catalog name
- `catalog_type` - "duckdb", "sqlite", or "postgres"
- `metadata_path` - path to catalog file/connection string

```{r}
.db_get_con <- function() {
  # Returns the connection if it exists and is valid, NULL otherwise
  if (exists("con", envir = .db_env) && DBI::dbIsValid(.db_env$con)) {
    return(.db_env$con)
  }
  NULL
}
```

This is the safe way to get the connection - it checks validity, not just existence.

```{r}
.db_get <- function(name, default = NULL) {
  # Safe getter with default value
  if (exists(name, envir = .db_env)) {
    get(name, envir = .db_env)
  } else {
    default
  }
}
```

Used throughout the package to retrieve stored values.

### Catalog Backend Helpers

These internal functions handle the different catalog backend types:

```{r}
.db_build_ducklake_dsn <- function(catalog_type, metadata_path) {
  switch(catalog_type,
    duckdb = paste0("ducklake:", metadata_path),
    sqlite = paste0("ducklake:sqlite:", metadata_path),
    postgres = paste0("ducklake:postgres:", metadata_path),
    stop("Unknown catalog_type")
  )
}
```

This builds the correct DuckLake connection string based on the backend:
- DuckDB: `ducklake:metadata.ducklake`
- SQLite: `ducklake:sqlite:catalog.sqlite`
- PostgreSQL: `ducklake:postgres:dbname=... host=...`

```{r}
.db_load_catalog_extensions <- function(con, catalog_type) {
  # Always need ducklake
  try(DBI::dbExecute(con, "INSTALL ducklake"), silent = TRUE)
  DBI::dbExecute(con, "LOAD ducklake")

  # Load backend-specific extension
  if (catalog_type == "sqlite") {
    try(DBI::dbExecute(con, "INSTALL sqlite"), silent = TRUE)
    DBI::dbExecute(con, "LOAD sqlite")
  } else if (catalog_type == "postgres") {
    try(DBI::dbExecute(con, "INSTALL postgres"), silent = TRUE)
    DBI::dbExecute(con, "LOAD postgres")
  }
}
```

DuckDB needs extra extensions loaded for SQLite and PostgreSQL backends.

### `db_connect()` - DuckLake Connection

```{r}
db_connect <- function(duckdb_db = ":memory:",
                       catalog = "cso",
                       catalog_type = c("duckdb", "sqlite", "postgres"),
                       metadata_path = "metadata.ducklake",
                       data_path = "/data/lake",
                       snapshot_version = NULL,
                       snapshot_time = NULL,
                       ...) {

  catalog_type <- match.arg(catalog_type)

  # Create DuckDB connection
  con <- DBI::dbConnect(duckdb::duckdb(), dbdir = duckdb_db)

  # Load DuckLake and backend-specific extensions
  .db_load_catalog_extensions(con, catalog_type)

  # Build the backend-specific connection string
  ducklake_dsn <- .db_build_ducklake_dsn(catalog_type, metadata_path)

  # Build ATTACH statement with options
  attach_opts <- c(glue::glue("DATA_PATH {.db_sql_quote(data_path)}"))

  if (!is.null(snapshot_version)) {
    attach_opts <- c(attach_opts, glue::glue("SNAPSHOT_VERSION {as.integer(snapshot_version)}"))
  }

  attach_sql <- glue::glue(
    "ATTACH {.db_sql_quote(ducklake_dsn)} AS {catalog} ({paste(attach_opts, collapse = ', ')})"
  )

  # Attach with helpful error messages
  tryCatch({
    DBI::dbExecute(con, attach_sql)
  }, error = function(e) {
    DBI::dbDisconnect(con, shutdown = TRUE)
    hint <- switch(catalog_type,
      sqlite = "Ensure the sqlite extension is available and the metadata file path is accessible.",
      postgres = "Ensure PostgreSQL is running and the connection string is correct.",
      duckdb = "Ensure the metadata file path is accessible."
    )
    stop("Failed to attach DuckLake catalog.\n", hint, "\n\nOriginal error: ", e$message)
  })

  DBI::dbExecute(con, glue::glue("USE {catalog}"))

  # Store connection info
  assign("con", con, envir = .db_env)
  assign("catalog", catalog, envir = .db_env)
  assign("catalog_type", catalog_type, envir = .db_env)
  assign("metadata_path", metadata_path, envir = .db_env)
  assign("data_path", data_path, envir = .db_env)
  ...
}
```

**Key points:**

- Supports three catalog backends (DuckDB, SQLite, PostgreSQL)
- ATTACH statement connects the DuckLake catalog to the DuckDB session
- Error messages include backend-specific troubleshooting hints

---

## `R/db_write.R` - Writing Data

### `db_write()` - DuckLake Tables

`db_write()` supports two input types:

1. **data.frame/tibble** - Data passes through R memory
2. **Lazy dbplyr table** - Data stays entirely in DuckDB (zero-copy)

```{r}
db_write <- function(data, schema = "main", table,
                     mode = c("overwrite", "append"),
                     col_types = NULL,       # Explicit column types (data.frame only)
                     partition_by = NULL,    # Hive partitioning
                     bucket_by = NULL,       # Bucket partitioning
                     sort_by = NULL,         # Clustering
                     inline = FALSE,         # Data inlining
                     commit_author = NULL,
                     commit_message = NULL) {

  # Detect if input is a lazy dbplyr table
  is_lazy <- inherits(data, "tbl_lazy") || inherits(data, "tbl_sql")

  if (is_lazy) {
    # ===== LAZY TABLE PATH =====
    # Extract SQL from the dbplyr query
    select_sql <- dbplyr::sql_render(data)

    # Execute directly in DuckDB - no R memory used
    if (mode == "overwrite") {
      DBI::dbExecute(con, glue("CREATE TABLE {qname} AS {select_sql}"))
    } else {
      DBI::dbExecute(con, glue("INSERT INTO {qname} {select_sql}"))
    }

  } else {
    # ===== DATA.FRAME PATH =====
    # Register data temporarily in DuckDB
    tmp <- .db_temp_name()
    duckdb::duckdb_register(con, tmp, data)

    # Build column definitions with explicit types
    cols <- vapply(data, .db_r_to_duckdb_type, character(1))
    if (!is.null(col_types)) {
      for (col_name in names(col_types)) {
        cols[col_name] <- col_types[[col_name]]
      }
    }

    # Create table and insert from temp view
    DBI::dbExecute(con, glue("CREATE TABLE {qname} ({col_defs})"))
    DBI::dbExecute(con, glue("INSERT INTO {qname} SELECT * FROM {tmp}"))
  }

  # Both paths: apply partitioning, clustering, etc.
}
```

**Key points:**

- Uses transactions (`BEGIN`/`COMMIT`) for atomicity
- `ROLLBACK` on error ensures no partial changes
- Lazy tables use `dbplyr::sql_render()` to extract SQL, then `CREATE TABLE AS SELECT`
- `col_types` only applies to data.frame input (lazy tables get types from query)
- Supports hive partitioning, bucket partitioning, clustering, and data inlining

**Example: Zero-copy pipeline**

```{r}
# Data never leaves DuckDB - efficient for in-lake transforms
db_read(table = "raw_events") |>
  filter(year == 2024) |>
  group_by(event_type) |>
  summarise(count = n()) |>
  db_write(table = "event_counts")
# Message: Wrote data to cso.main.event_counts (from query)
```

### `R/arrow.R` - Arrow Integration

```{r}
# Read directly as Arrow Table (bypasses DuckDB compute)
db_read_arrow <- function(schema = "main", table, columns = NULL, as_data_frame = TRUE) {
  # Get parquet file paths from DuckLake metadata
  files <- .db_get_table_files(con, catalog, schema, table)

  # Read via Arrow
  if (length(files) == 1) {
    arrow::read_parquet(files[1], col_select = columns)
  } else {
    arrow::open_dataset(files) |> dplyr::collect()
  }
}

# Write Arrow Table to DuckLake
db_write_arrow <- function(data, schema = "main", table, mode = c("overwrite", "append"), ...) {
  # Convert Arrow to data.frame and delegate to db_write
  if (inherits(data, c("ArrowTabular", "Table"))) {
    data <- as.data.frame(data)
  }
  db_write(data, schema, table, mode, ...)
}
```

**When to use Arrow functions:**

- `db_read_arrow()`: Direct Parquet access, Arrow ecosystem interop, large data transfers
- `db_write_arrow()`: When data is already in Arrow format

---

## `R/upsert.R` - MERGE Operations

### `db_upsert()` - Update + Insert

```{r}
db_upsert <- function(data, schema = "main", table, by,
                      strict = TRUE, update_cols = NULL,
                      commit_author = NULL, commit_message = NULL) {

  # Validate key columns exist in both data and target
  missing_keys <- setdiff(by, names(data))
  if (length(missing_keys) > 0) stop(...)

  # Strict mode: reject duplicate keys in incoming data
  if (strict) {
    dup_check_sql <- glue::glue("
      SELECT {by_sql}, COUNT(*) AS n
      FROM {tmp}
      GROUP BY {by_sql}
      HAVING COUNT(*) > 1
    ")
    dups <- DBI::dbGetQuery(con, dup_check_sql)
    if (nrow(dups) > 0) stop("Duplicate keys found...")
  }

  # Build MERGE statement
  on_sql <- paste(glue::glue("t.{by} = s.{by}"), collapse = " AND ")

  # UPDATE clause depends on update_cols
  update_clause <- if (is.null(update_cols)) {
    "WHEN MATCHED THEN UPDATE"  # DuckDB shorthand: update all
  } else if (length(update_cols) == 0) {
    ""  # Insert-only mode
  } else {
    glue::glue("WHEN MATCHED THEN UPDATE SET {update_sql}")
  }

  merge_sql <- glue::glue("
    MERGE INTO {qname} AS t
    USING {tmp} AS s
    ON ({on_sql})
    {update_clause}
    WHEN NOT MATCHED THEN INSERT ...
  ")

  # Execute in transaction
  DBI::dbExecute(con, "BEGIN")
  DBI::dbExecute(con, merge_sql)
  DBI::dbExecute(con, "COMMIT")
}
```

**Key points:**

- `MERGE INTO` is standard SQL for upsert operations
- `strict = TRUE` prevents accidental duplicates (a common data quality issue)
- `update_cols = NULL` means update everything; `character(0)` means insert-only
- All wrapped in a transaction for safety

---

## `R/docs.R` - Documentation & Data Dictionary

### Metadata Storage

Documentation metadata is stored using DuckLake's **native COMMENT ON** with JSON for structured data. This approach:

- Survives disconnect/reconnect (unlike custom metadata tables)
- Supports time travel (comments are versioned with snapshots)
- Is stored in the DuckLake catalog alongside table definitions

```{r}
# Internal helper to parse JSON comment
.db_parse_comment <- function(comment_str) {
  if (is.null(comment_str) || !nzchar(comment_str)) return(NULL)

  tryCatch({
    parsed <- jsonlite::fromJSON(comment_str)
    if (is.list(parsed)) return(parsed)
    list(description = comment_str)  # Plain string
  }, error = function(e) {
    list(description = comment_str)  # Not JSON, treat as description
  })
}

# Get table comment from DuckDB catalog
.db_get_table_comment <- function(con, catalog, schema, table) {
  result <- DBI::dbGetQuery(con, glue::glue("
    SELECT comment FROM duckdb_tables()
    WHERE database_name = ... AND schema_name = ... AND table_name = ...
  "))
  if (nrow(result) == 0 || is.na(result$comment[1])) return(NULL)
  result$comment[1]
}
```

### `db_get_docs()` - Retrieve Documentation

```{r}
db_get_docs <- function(schema = "main", table) {
  # Read table comment from duckdb_tables()
  table_comment <- .db_get_table_comment(con, catalog, schema, table)
  parsed <- .db_parse_comment(table_comment)

  # Build result with defaults

  result <- list(
    schema = schema, table = table,
    description = parsed$description,
    owner = parsed$owner,
    tags = parsed$tags %||% character(0),
    lineage = if (!is.null(parsed$lineage_sources)) list(
      sources = parsed$lineage_sources,
      transformation = parsed$lineage_transformation
    ),
    columns = list()  # Populated from duckdb_columns()
  )

  # Get column comments
  cols <- DBI::dbGetQuery(con, "SELECT column_name, comment FROM duckdb_columns() ...")
  for (col in cols with comments) {
    result$columns[[col_name]] <- .db_parse_comment(col_comment)
  }

  result
}
```

### `db_dictionary()` - Generate Data Dictionary

```{r}
db_dictionary <- function(schema = NULL, include_columns = TRUE) {
  # Query information_schema for tables
  tables <- DBI::dbGetQuery(con, "SELECT * FROM information_schema.tables ...")

  for (tbl in tables) {
    # Get docs from native comments via db_get_docs()
    docs <- db_get_docs(schema = ..., table = ...)
    cols <- DBI::dbGetQuery(con, "SELECT * FROM information_schema.columns ...")
  }

  do.call(rbind, rows)
}
```

### `db_search()` - Find Tables

```{r}
db_search <- function(pattern, field = c("all", "name", "description", "owner", "tags")) {
  # Get full dictionary (without columns for speed)
  dict <- db_dictionary(include_columns = FALSE)

  # Filter by pattern match
  matches <- switch(field,
    all = grepl(pattern, dict$name) | grepl(pattern, dict$description) | ...,
    name = grepl(pattern, dict$name),
    description = grepl(pattern, dict$description),
    ...
  )

  dict[matches, ]
}
```

### `db_lineage()` - Record Data Lineage

Lineage is stored in the table's native COMMENT as JSON alongside other metadata:

```{r}
db_lineage <- function(schema = "main", table, sources, transformation = NULL) {
  # Get existing table comment/metadata
  existing_comment <- .db_get_table_comment(con, catalog, schema, table)
  existing_meta <- .db_parse_comment(existing_comment) %||% list()

  # Add lineage info (preserves existing description, owner, tags)
  existing_meta$lineage_sources <- sources
  existing_meta$lineage_transformation <- transformation

  # Save back using db_comment
  db_comment(schema = schema, table = table, comment = existing_meta)
}
```

**Key points:**

- All metadata stored in native COMMENT ON (survives disconnect/reconnect)
- JSON format allows structured data (description, owner, tags, lineage)
- `db_get_lineage()` extracts lineage from the parsed comment JSON
- Lineage is merged with existing metadata, not overwritten

---

## `R/preview.R` - Write Previews

### `db_preview_write()` - Preview Before Writing

```{r}
db_preview_write <- function(data, schema = "main", table,
                             mode = c("overwrite", "append")) {

  preview <- list(
    mode = mode,
    table_exists = db_table_exists(schema = schema, table = table),
    incoming = list(
      rows = nrow(data),
      cols = ncol(data),
      columns = names(data)
    )
  )

  # If target exists, compare schemas
  if (preview$table_exists) {
    existing_cols <- db_table_cols(schema = schema, table = table)

    preview$schema_changes <- list(
      new_columns = setdiff(names(data), existing_cols),
      removed_columns = setdiff(existing_cols, names(data))
    )
  }

  .print_preview(preview)
  invisible(preview)
}
```

**Key points:**

- Gathers all relevant information without making any changes
- Compares incoming schema with existing schema
- Returns structured data for programmatic use, prints human-readable summary

### `db_preview_upsert()` - Preview Insert vs Update Counts

```{r}
db_preview_upsert <- function(data, schema, table, by, update_cols = NULL) {

  # Register data temporarily
  tmp <- .db_temp_name()
  duckdb::duckdb_register(con, tmp, data)

  # Count matches (updates)
  match_sql <- glue::glue("
    SELECT COUNT(*) FROM {tmp} s
    INNER JOIN {qname} t ON {join_on_keys}
  ")
  matches <- DBI::dbGetQuery(con, match_sql)$n

  # Check for duplicate keys
  dup_sql <- glue::glue("
    SELECT {key_cols}, COUNT(*) FROM {tmp}
    GROUP BY {key_cols} HAVING COUNT(*) > 1
  ")
  dups <- DBI::dbGetQuery(con, dup_sql)

  preview$impact <- list(
    inserts = nrow(data) - matches,
    updates = matches,
    duplicates_in_incoming = nrow(dups)
  )

  .print_upsert_preview(preview)
  invisible(preview)
}
```

---

## `R/discovery.R` - Finding Data

### DuckLake Discovery

```{r}
db_tables <- function(schema = "main") {
  sql <- glue::glue("
    SELECT table_name
    FROM information_schema.tables
    WHERE table_catalog = {.db_sql_quote(catalog)}
      AND table_schema  = {.db_sql_quote(schema)}
      AND table_type    = 'BASE TABLE'
  ")
  DBI::dbGetQuery(con, sql)$table_name
}
```

**Key points:**

- Uses standard `information_schema` views
- Filters by catalog to avoid confusion with other attached databases

### Views and Macros

Views and macros are stored in the DuckLake catalog and support time travel:

```{r}
# Create a view - stored in catalog
db_create_view(view = "active_users",
               query = "SELECT * FROM users WHERE active = true")

# Create a scalar macro
db_create_macro(name = "add_vals", params = c("a", "b"), body = "a + b")

# Create a table macro (returns a table, not a scalar)
db_create_macro(name = "big_sales",
                params = c(threshold = "INTEGER"),
                body = "SELECT * FROM sales WHERE value > threshold",
                table_macro = TRUE)

# List and drop
db_list_views()
db_list_macros()
db_drop_view(view = "active_users")
db_drop_macro(name = "add_vals")
```

### Documentation with db_comment()

`db_comment()` accepts either a **list** (stored as JSON) or a **string** (stored as-is):

```{r}
# Structured metadata (stored as JSON)
db_comment(table = "users", comment = list(
  description = "Active user accounts",
  owner = "Admin Team",
  tags = c("users", "reference")
))

# Simple string comment
db_comment(table = "users", comment = "Active user accounts")

# Column documentation
db_comment(table = "users", column = "email", comment = list(
  description = "Primary email",
  units = NULL
))

# Remove comment
db_comment(table = "users", column = "email", comment = NULL)
```

The list format is automatically converted to JSON via `jsonlite::toJSON()`. When retrieved via `db_get_docs()`, JSON is parsed back to a list.

### Logging

Enable DuckDB query logging for debugging:

```{r}
db_enable_logging(TRUE, log_type = "query")
# ... run operations ...
db_query("SELECT * FROM duckdb_logs()")
db_enable_logging(FALSE)
```

---

## `R/maintenance.R` - Admin Operations

### `db_vacuum()` - Clean Up Old Snapshots

```{r}
db_vacuum <- function(older_than = "30 days", dry_run = TRUE) {
  # Get snapshots before
  snapshots_before <- db_snapshots()

  if (dry_run) {
    # Calculate what would be removed
    # Show preview
    return(invisible(to_remove))
  }

  # Actually vacuum
  DBI::dbExecute(con, glue::glue(
    "CALL ducklake_vacuum({.db_sql_quote(catalog)}, INTERVAL '30 days')"
  ))
}
```

**Key points:**

- `dry_run = TRUE` by default prevents accidental deletion
- Shows clear before/after summary

### `db_rollback()` - Restore Previous Version

```{r}
db_rollback <- function(schema, table, version = NULL, timestamp = NULL) {
  # Build time travel clause
  at_clause <- glue::glue("AT (VERSION => {version})")

  # Rollback = create new version with old data
  rollback_sql <- glue::glue(
    "CREATE OR REPLACE TABLE {qname} AS SELECT * FROM {qname} {at_clause}"
  )

  DBI::dbExecute(con, rollback_sql)
}
```

**Key points:**

- Rollback creates a NEW snapshot with the old data (non-destructive)
- You can always "roll forward" by rolling back to a later version

### `db_diff()` - Compare Versions

```{r}
db_diff <- function(schema, table, from_version, to_version, key_cols = NULL) {
  # ADDED: rows in 'to' but not in 'from'
  added_sql <- glue::glue("SELECT * FROM {to_ref} EXCEPT SELECT * FROM {from_ref}")

  # REMOVED: rows in 'from' but not in 'to'
  removed_sql <- glue::glue("SELECT * FROM {from_ref} EXCEPT SELECT * FROM {to_ref}")

  result <- list(
    added = DBI::dbGetQuery(con, added_sql),
    removed = DBI::dbGetQuery(con, removed_sql)
  )

  # If key_cols provided, find modified rows (same key, different values)
  if (!is.null(key_cols)) {
    # Modified = key appears in both added and removed
    modified_sql <- glue::glue("
      SELECT new_tbl.*
      FROM ({added_sql}) AS new_tbl
      INNER JOIN ({removed_sql}) AS old_tbl
      ON {join_on_keys}
    ")
    result$modified <- DBI::dbGetQuery(con, modified_sql)
  }

  result
}
```

### `db_compact()` - Merge Small Files

```{r}
db_compact <- function(schema = "main", table = NULL, max_files = NULL) {
  # Build the SQL call with optional parameters
  sql <- glue::glue("CALL ducklake_merge_adjacent_files({catalog}, {table}, ...)")

  # Get file stats before and after for reporting
  stats_before <- db_file_stats(schema, table)
  DBI::dbExecute(con, sql)
  stats_after <- db_file_stats(schema, table)

  # Report reduction
  message("Files before: ", sum(stats_before$file_count))
  message("Files after:  ", sum(stats_after$file_count))
}
```

**Key points:**

- Uses DuckLake's `ducklake_merge_adjacent_files()` procedure
- `max_files` parameter limits memory usage for large tables
- Files with incompatible schema versions cannot be merged
- Run `db_cleanup_files()` after to remove old files

### `db_file_stats()` - Check File Statistics

```{r}
db_file_stats <- function(schema = NULL, table = NULL) {
  # Query DuckLake metadata tables directly
  # Tables are in __ducklake_metadata_{catalog} schema
  meta_schema <- glue::glue("__ducklake_metadata_{catalog}")

  sql <- glue::glue("
    SELECT
      s.schema_name, t.table_name,
      COUNT(f.data_file_id) AS file_count,
      COALESCE(SUM(f.record_count), 0) AS total_rows,
      COALESCE(SUM(f.file_size_bytes), 0) AS total_bytes
    FROM {meta_schema}.ducklake_table t
    JOIN {meta_schema}.ducklake_schema s ON t.schema_id = s.schema_id
    LEFT JOIN {meta_schema}.ducklake_data_file f
      ON t.table_id = f.table_id AND f.end_snapshot IS NULL
    WHERE t.end_snapshot IS NULL AND s.end_snapshot IS NULL
    GROUP BY s.schema_name, t.table_name
  ")
  info <- DBI::dbGetQuery(con, sql)

  info$avg_file_bytes <- ifelse(info$file_count > 0, info$total_bytes / info$file_count, 0)
  info$avg_rows_per_file <- ifelse(info$file_count > 0, info$total_rows / info$file_count, 0)
  info
}
```

**Key points:**

- Queries DuckLake metadata tables directly (`ducklake_table`, `ducklake_schema`, `ducklake_data_file`)
- Filters by `end_snapshot IS NULL` to get only current (not deleted) data
- High file count with small average size indicates compaction needed
- Useful for monitoring and maintenance scheduling

### `db_cleanup_files()` - Remove Orphaned Files

```{r}
db_cleanup_files <- function(dry_run = TRUE) {
  if (dry_run) {
    # Show preview of what would be cleaned
    return(invisible(NA))
  }

  sql <- glue::glue("CALL ducklake_cleanup_old_files({.db_sql_quote(catalog)})")
  DBI::dbExecute(con, sql)
}
```

**Key points:**

- Orphaned files are created by `db_vacuum()` and `db_compact()`
- `dry_run = TRUE` by default for safety
- Reclaims disk space by removing unreferenced Parquet files

---

## `R/advanced.R` - Advanced Features

This file contains advanced DuckLake features: clustering, inlining, and Iceberg export.

### `db_set_clustering()` - Configure Sort Order

```{r}
db_set_clustering <- function(schema = "main", table, columns) {
  qname <- paste0(catalog, ".", schema, ".", table)

  if (is.null(columns) || length(columns) == 0) {
    sql <- glue::glue("ALTER TABLE {qname} RESET SORTED BY")
  } else {
    sort_clause <- paste(columns, collapse = ", ")
    sql <- glue::glue("ALTER TABLE {qname} SET SORTED BY ({sort_clause})")
  }

  DBI::dbExecute(con, sql)
}
```

**Key points:**

- Uses DuckLake's `ALTER TABLE ... SET SORTED BY` to configure sort order
- Clustering improves performance for range scans on sorted columns
- Only affects new data - use `db_recluster()` for existing data

### `db_recluster()` - Re-sort Existing Data

```{r}
db_recluster <- function(schema = "main", table, max_files = NULL) {
  # DuckLake sorts data during compaction based on SET SORTED BY order
  # Use ducklake_merge_adjacent_files to compact and re-sort
  args <- c(.db_sql_quote(catalog), .db_sql_quote(table))
  named_args <- c(glue::glue("schema => {.db_sql_quote(schema)}"))

  if (!is.null(max_files)) {
    named_args <- c(named_args, glue::glue("max_compacted_files => {as.integer(max_files)}"))
  }

  all_args <- c(args, named_args)
  sql <- glue::glue("CALL ducklake_merge_adjacent_files({paste(all_args, collapse = ', ')})")

  DBI::dbExecute(con, sql)
}
```

**Key points:**

- Uses `ducklake_merge_adjacent_files()` to compact and re-sort data
- DuckLake automatically applies the table's sort order during compaction
- Note the `=>` syntax for named parameters (standard for DuckLake functions)

### `db_set_inline_threshold()` - Configure Data Inlining

```{r}
db_set_inline_threshold <- function(schema = "main", table = NULL, threshold = 10) {
  # Build set_option call using := syntax for named parameters
  args <- c("'data_inlining_row_limit'", as.integer(threshold))

  if (!is.null(table)) {
    args <- c(args, glue::glue("table_name := {.db_sql_quote(table)}"))
  }
  if (!is.null(schema)) {
    args <- c(args, glue::glue("schema := {.db_sql_quote(schema)}"))
  }

  sql <- glue::glue("CALL {catalog}.set_option({paste(args, collapse = ', ')})")
  DBI::dbExecute(con, sql)
}
```

**Key points:**

- Uses catalog's `set_option()` procedure to configure inlining threshold
- Note the `:=` syntax for named parameters (specific to set_option)
- Inlining stores small writes in the catalog instead of creating parquet files
- Flush with `db_flush_inlined()` when ready to write to parquet

---

## `R/zzz.R` - Package Hooks

```{r}
# Null coalesce operator used throughout
`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

.onAttach <- function(libname, pkgname) {
  packageStartupMessage(
    "datapond ", utils::packageVersion("datapond"), "\n",
    "Use db_connect() to connect to a DuckLake catalog."
  )
}

.onUnload <- function(libpath) {
  # Clean up connection when package unloads
  con <- .db_get_con()
  if (!is.null(con)) {
    try(DBI::dbDisconnect(con, shutdown = TRUE), silent = TRUE)
  }
}
```

**Key points:**

- `.onAttach` runs when the package is loaded (shows helpful message)
- `.onUnload` ensures cleanup if the package is unloaded
- `%||%` handles both `NULL` and zero-length values

---

## `R/browser.R`, `R/browser_ui.R`, `R/browser_server.R` - Interactive Browser

These files implement a Shiny gadget for interactively browsing the data lake.

### Architecture

The browser uses the **Shiny module** pattern, which allows it to be:
1. Used standalone via `db_browser()`
2. Embedded in other Shiny apps via `db_browser_ui()` + `db_browser_server()`

```{r}
# Standalone usage
db_browser()

# Embedded in another app
ui <- fluidPage(
  db_browser_ui("my_browser")
)
server <- function(input, output, session) {
  db_browser_server("my_browser")
}
```

### `browser.R` - Launcher

```{r}
db_browser <- function(height = "500px", viewer = c("dialog", "browser", "pane")) {
  # Check packages are available
  .db_assert_browser_packages()

  # Build app
  ui <- db_browser_app_ui(height = height)
  server <- function(input, output, session) {
    db_browser_server(id = "db_browser", height = height)
  }
  app <- shiny::shinyApp(ui = ui, server = server)

  # Run as gadget
  shiny::runGadget(app, viewer = viewer_func, stopOnCancel = TRUE)
}
```

### `browser_ui.R` - Module UI

The UI is built with `bslib` for modern Bootstrap 5 styling:

```{r}
db_browser_ui <- function(id, height = "500px") {
  ns <- shiny::NS(id)  # Namespace for module

  # Sidebar with tree view
  sidebar <- bslib::sidebar(
    shiny::uiOutput(ns("tree_view")),
    shiny::actionButton(ns("refresh_tree"), "Refresh")
  )

  # Main content with tabs
  main_content <- bslib::navset_card_tab(
    bslib::nav_panel("Preview", ...),
    bslib::nav_panel("Metadata", ...),
    bslib::nav_panel("Search", ...),
    bslib::nav_panel("Dictionary", ...)
  )

  bslib::page_sidebar(sidebar = sidebar, main_content)
}
```

**Key points:**

- `NS(id)` creates namespaced IDs so multiple instances don't conflict
- `bslib` provides modern Bootstrap 5 components
- Tree view is rendered dynamically based on available schemas and tables

### `browser_server.R` - Module Server

```{r}
db_browser_server <- function(id, height = "500px") {
  shiny::moduleServer(id, function(input, output, session) {

    # Reactive values for selection state
    rv <- shiny::reactiveValues(
      selected_schema = NULL,
      selected_table = NULL
    )

    # Tree view
    output$tree_view <- shiny::renderUI({
      .render_ducklake_tree(ns, rv)
    })

    # Preview - loads data on button click
    preview_data <- shiny::eventReactive(input$load_preview, {
      db_read(rv$selected_schema, rv$selected_table) |>
        head(input$preview_rows) |>
        collect()
    })

    output$preview_table <- DT::renderDataTable({
      DT::datatable(preview_data(), filter = "top")
    })

    # Search
    search_results <- shiny::eventReactive(input$do_search, {
      db_search(input$search_pattern, field = input$search_field)
    })

    # Dictionary with download
    output$download_dict <- shiny::downloadHandler(
      filename = "data_dictionary.csv",
      content = function(file) {
        write.csv(db_dictionary(), file)
      }
    )
  })
}
```

**Key points:**

- `moduleServer()` creates the namespaced server logic
- `reactiveValues` tracks UI state (selected table, etc.)
- `eventReactive` triggers expensive operations only on button click
- `DT::datatable` provides interactive tables with filtering

---

## Security Considerations

### SQL Injection Prevention

All user inputs that go into SQL are either:

1. **Validated** with `.db_validate_name()` (only allows `A-Za-z0-9_-`)
2. **Quoted** with `.db_sql_quote()` (escapes single quotes)

```{r}
# This is safe:
table <- .db_validate_name(table)  # Rejects "../../../etc"
path <- .db_sql_quote(glob_path)   # Escapes quotes properly
```

### Path Traversal Prevention

`.db_validate_name()` rejects any input containing:
- Path separators (`/`, `\`)
- Parent directory references (`..`)
- Special characters

This prevents attacks like:
```{r}
db_read(table = "../../../etc/passwd")  # Rejected!
```

---

## Testing Your Understanding

Try answering these questions:

1. Why do we use `.db_get_con()` instead of accessing `.db_env$con` directly?
2. What happens if `db_write()` fails halfway through?
3. Why does `db_read()` return a lazy table instead of collected data?
4. What's the difference between `update_cols = NULL` and `update_cols = character(0)` in `db_upsert()`?
5. Why would you choose SQLite over DuckDB as a catalog backend?
6. How does `db_preview_upsert()` know how many rows will be updated vs inserted?
7. Where is documentation metadata stored? (Hint: native COMMENT ON with JSON)

---

## Contributing to the Package

When adding new functions:

1. **Always validate inputs** with `.db_validate_name()` or custom validation
2. **Check connection** at the start of every function
3. **Use `.db_get()` and `.db_get_con()`** for accessing stored state
4. **Return invisibly** for side-effect functions, visibly for queries
5. **Add roxygen documentation** with `@param`, `@return`, `@examples`, `@export`
6. **Update NAMESPACE** if adding new exports
7. **Add tests** in `tests/testthat/` for new functionality

Run `devtools::document()` after adding roxygen comments to regenerate NAMESPACE and help files.
