---
title: "Data Lake Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Lake Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette explains the core concepts behind `datapond` - what data lakes are, how hive partitioning works, and what DuckLake brings to the table. No prior knowledge assumed.

## The Problem We're Solving

Within an organisation, different sections need to share data:

- **Trade** produces import/export figures that **National Accounts** needs
- **Labour** produces employment data that multiple sections consume
- **Health** produces statistics that feed into other analyses

Currently, this happens through shared network drives with folder-based access control. It works, but has limitations:

- No versioning (if someone overwrites a file, the old version is gone)
- No way to know when data changed or who changed it
- Large files are slow to query (must read the whole thing)
- No schema enforcement (columns can change without warning)

`datapond` addresses these issues while keeping the familiar folder-and-permissions model that IT already supports.

---

## What is a Data Lake?

A **data lake** is a storage system designed to hold large amounts of structured data in files (typically Parquet format) rather than in a traditional database.

```
Traditional Database          Data Lake
┌─────────────────┐          ┌─────────────────┐
│  Database       │          │  Files on disk  │
│  Server         │          │  (Parquet)      │
│                 │          │                 │
│  ┌───────────┐  │          │  📁 Trade/      │
│  │ Table A   │  │          │    📄 data.parquet
│  ├───────────┤  │          │  📁 Labour/     │
│  │ Table B   │  │          │    📄 data.parquet
│  └───────────┘  │          │                 │
└─────────────────┘          └─────────────────┘
     ↓                              ↓
  Needs server               Just files!
  Needs DBA                  Query with DuckDB
  Licensing costs            Free & fast
```

**Why Parquet?**

Parquet is a columnar file format that's:

- **Compressed** - files are 5-10x smaller than CSV
- **Fast** - only reads the columns you need
- **Typed** - preserves data types (dates, numbers, strings)
- **Universal** - works with R, Python, SAS, Excel, and more

---

## Hive Partitioning Explained

**Hive partitioning** is a way of organising files into folders based on column values. The folder names encode the data they contain.

### Example: Trade Imports Data

Instead of one massive file:
```
Trade/Imports/all_data.parquet  (10 GB - slow to read!)
```

We partition by year and month:
```
Trade/Imports/
  year=2023/
    month=01/
      data.parquet
      data_part2.parquet
    month=02/
      data.parquet
    ...
  year=2024/
    month=01/
      data.parquet
    ...
```

### Why This Matters

When you query for January 2024 data:

```{r}
imports |> 
  filter(year == 2024, month == 1)
```

DuckDB is smart enough to **only read the `year=2024/month=01/` folder**. It skips all other folders entirely. This is called **partition pruning** and it makes queries on large datasets dramatically faster.

### Choosing Partition Columns

Good partition columns are:

- **Low cardinality** - not too many unique values (year, month, region)
- **Frequently filtered** - columns you often use in WHERE clauses
- **Stable** - values that don't change for existing records

Bad partition columns:

- **High cardinality** - unique IDs, timestamps with seconds
- **Rarely filtered** - columns you never filter on

**Rule of thumb**: Aim for partition folders containing 100MB - 1GB of data each.

### How datapond Uses Hive Partitioning

```{r}
# Write with partitioning
db_hive_write(
  my_data,
  section = "Trade",
  dataset = "Imports",
  partition_by = c("year", "month")  # Creates year=X/month=Y folders
)

# Read - partitions are automatic columns
imports <- db_hive_read("Trade", "Imports")
imports |> filter(year == 2024)  # Only reads 2024 folders
```

---

## What is DuckLake?

**DuckLake** is a newer approach that adds database-like features on top of Parquet files. Think of it as "Parquet files with superpowers".

```
Hive Partitioning              DuckLake
┌─────────────────┐          ┌─────────────────┐
│ Just files      │          │ Files +         │
│                 │          │ Metadata catalog│
│ 📁 year=2024/   │          │                 │
│   📄 data.parquet          │ 📄 catalog.sqlite
│                 │          │ 📁 data/        │
│ No versioning   │          │   📄 file1.parquet
│ No transactions │          │   📄 file2.parquet
│                 │          │                 │
└─────────────────┘          │ ✅ Versioning   │
                             │ ✅ Transactions │
                             │ ✅ Time travel  │
                             └─────────────────┘
```

### The Metadata Catalog

DuckLake needs a place to store metadata about your tables. This **catalog** tracks:

- Which Parquet files belong to which table
- The schema (columns and types) of each table
- A history of all changes (snapshots)
- Who made changes and when

The actual data is still in Parquet files - DuckLake just adds a management layer.

---

## Choosing a Catalog Backend

DuckLake supports three different backends for storing this catalog metadata:

### 1. DuckDB (Single User)

```{r}
db_lake_connect(
  catalog_type = "duckdb",
  metadata_path = "metadata.ducklake",
  data_path = "//CSO-NAS/DataLake/data"
)
```

- Metadata stored in a `.ducklake` file
- **Single client only** - if two people connect, one will fail
- Good for: personal use, development, testing

### 2. SQLite (Multiple Local Users) - RECOMMENDED

```{r}
db_lake_connect(
  catalog_type = "sqlite",
  metadata_path = "//CSO-NAS/DataLake/catalog.sqlite",
  data_path = "//CSO-NAS/DataLake/data"
)
```

- Metadata stored in a `.sqlite` file on the network drive
- **Multiple readers + single writer** with automatic retry
- Still just a file - no server needed
- Good for: **shared network drives, most use cases**

**How SQLite handles concurrency:**

When someone is writing, other writers will wait and retry automatically. Readers can continue uninterrupted. This works well for typical usage where writes are less frequent than reads.

### 3. PostgreSQL (Multi-User Lakehouse)

```{r}
db_lake_connect(
  catalog_type = "postgres",
  metadata_path = "dbname=ducklake_catalog host=db.cso.ie",
  data_path = "//CSO-NAS/DataLake/data"
)
```

- Metadata stored in a PostgreSQL database
- **Full concurrent access** - multiple readers and writers
- Requires PostgreSQL 12+ server
- Good for: large teams, high write concurrency, remote access

### Which Should You Choose?

```
Are you the only user?
  └─ Yes → DuckDB (simplest)
  └─ No → Are you on a shared network drive?
            └─ Yes → SQLite (recommended)
            └─ No → Do you need high write concurrency?
                      └─ Yes → PostgreSQL
                      └─ No → SQLite
```

**For most use cases, start with SQLite.** It's still just a file (familiar, works with IT permissions), but handles multiple users gracefully.

---

## Key DuckLake Features

### 1. Time Travel

Query data as it existed at any point in the past:

```{r}
# Current data
products <- db_lake_read(table = "products")

# Data as of version 5
products_v5 <- db_lake_read(table = "products", version = 5)

# Data as of last Tuesday
products_tue <- db_lake_read(table = "products", 
                              timestamp = "2025-01-14 00:00:00")
```

This is invaluable when:

- Someone reports "the numbers looked different yesterday" - you can check!
- You need to reproduce an analysis from a specific date
- Something went wrong and you need to see the before/after

### 2. Snapshots and Audit Trail

Every change creates a new **snapshot** with metadata:

```{r}
db_snapshots()
#>   snapshot_id snapshot_time       commit_author commit_message
#> 1           1 2025-01-01 09:00:00 jsmith        Initial load
#> 2           2 2025-01-15 14:30:00 mjones        Added Q4 data
#> 3           3 2025-01-20 11:00:00 mjones        Fixed country codes
```

You always know what changed, when, and (if recorded) why.

### 3. ACID Transactions

Changes are **atomic** - they either fully succeed or fully fail. No partial updates that leave data in a broken state.

```{r}
# If this fails halfway through, no data is changed
db_lake_write(big_dataset, table = "imports", mode = "overwrite")
```

### 4. Schema Evolution

DuckLake handles schema changes gracefully:

- Add a new column? Old data gets NULLs for that column
- Query old versions? They still work with the old schema

---

## Schemas: Organising Tables

In DuckLake, **schemas** are like folders for tables. They help organise related tables together.

```
catalog (datapond)
├── main (default schema)
│   └── reference_tables
├── trade
│   ├── imports
│   ├── exports
│   └── products
├── labour
│   ├── employment
│   └── earnings
└── health
    ├── hospitals
    └── waiting_times
```

### Using Schemas

```{r}
# Create a schema for your section
db_create_schema("trade")

# Write to it
db_lake_write(imports_data, schema = "trade", table = "imports")

# Read from it
imports <- db_lake_read(schema = "trade", table = "imports")

# List tables in a schema
db_tables("trade")
```

### Schemas vs Sections

| Hive Mode | DuckLake Mode |
|-----------|---------------|
| Section (folder) | Schema |
| Dataset (subfolder) | Table |
| `db_list_sections()` | `db_list_schemas()` |
| `db_list_datasets("Trade")` | `db_tables("trade")` |

They serve the same organisational purpose - grouping related data together.

---

## Data Documentation

Good data governance requires documentation. `datapond` provides built-in tools to document your datasets and generate a data dictionary.

### Documenting Datasets

```{r}
# Add metadata to a dataset
db_describe(
  section = "Trade",
  dataset = "Imports",
  description = "Monthly import values by country and HS commodity code",
  owner = "Trade Section",
  tags = c("trade", "monthly", "official")
)

# Document individual columns
db_describe_column(
  section = "Trade",
  dataset = "Imports",
  column = "value",
  description = "Import value",
  units = "EUR (thousands)"
)

db_describe_column(
  section = "Trade",
  dataset = "Imports",
  column = "country_code",
  description = "ISO 3166-1 alpha-2 country code"
)
```

### Where is Documentation Stored?

- **Hive mode**: Creates a `_metadata.json` file in each dataset folder
- **DuckLake mode**: Stores in a `_metadata` schema in the catalog

Either way, documentation travels with the data.

### Searching and Discovery

```{r}
# Search datasets by any field
db_search("trade")                        # Matches name, description, owner, or tags
db_search("monthly", field = "tags")      # Search only tags
db_search("Trade Section", field = "owner")  # Find datasets by owner

# Find columns across all datasets
db_search_columns("country")
#>   section  dataset  column_name  column_type  column_description
#> 1 Trade    Imports  country_code VARCHAR      ISO 3166-1 alpha-2...
#> 2 Trade    Exports  country_code VARCHAR      ISO 3166-1 alpha-2...
#> 3 Labour   Survey   country      VARCHAR      Country of residence
```

### Generating a Data Dictionary

```{r}
# Full data dictionary with column details
dict <- db_dictionary()

# Just dataset-level summary
dict_summary <- db_dictionary(include_columns = FALSE)

# Filter to specific section
dict_trade <- db_dictionary(section = "Trade")

# Export to Excel
writexl::write_xlsx(dict, "data_dictionary.xlsx")
```

The data dictionary includes:
- Dataset/table name and location
- Description, owner, tags
- Column names, types, and documentation
- Last updated timestamps

---

## Preview Before Writing

Before making changes to production data, you can preview what will happen:

### Hive Write Preview

```{r}
db_preview_hive_write(
  my_data,
  section = "Trade",
  dataset = "Imports",
  partition_by = c("year", "month"),
  mode = "replace_partitions"
)
```

Shows:
- Target path and whether it exists
- Incoming vs existing row counts
- **Schema changes** - new columns, removed columns, type changes
- **Partition impact** - which partitions will be affected

### DuckLake Write Preview

```{r}
db_preview_lake_write(my_data, table = "imports", mode = "overwrite")
```

Shows:
- Current vs new row counts
- Schema comparison
- Warnings (e.g., append to non-existent table)

### Upsert Preview

```{r}
db_preview_upsert(my_data, table = "products", by = "product_id")
```

Shows:
- How many rows will be **inserted** (new keys)
- How many rows will be **updated** (existing keys)
- Warnings about duplicate keys in incoming data

---

## Comparing the Two Approaches

### When to Use Hive Mode

✅ **Good for:**

- Migrating from existing SAS folder structure
- Simple data sharing without versioning needs
- When you need direct file access (for tools that can't use DuckLake)
- Getting started quickly with minimal setup

❌ **Limitations:**

- No versioning - overwrites are permanent
- No transaction safety - partial failures can corrupt data
- Manual schema management

### When to Use DuckLake Mode

✅ **Good for:**

- Production data pipelines where reliability matters
- When you need audit trails (who changed what, when)
- Analyses that need to be reproducible at specific points in time
- Complex updates (upserts, merges)

❌ **Considerations:**

- Slightly more complex setup (choose catalog backend)
- Newer technology (less battle-tested than simple files)

---

## Access Control

Both modes use **file system permissions** - the same model commonly used.

### Hive Mode Access

```
//CSO-NAS/DataLake/
├── Trade/          ← Trade section has read/write
│   ├── Imports/
│   └── Exports/
├── Labour/         ← Labour section has read/write
│   └── Employment/
└── Shared/         ← Everyone has read access
    └── Reference/
```

Users get access by being granted folder permissions through IT.

### DuckLake Mode Access

```
//CSO-NAS/DataLake/
├── catalog.sqlite      ← Need read (or read/write) access
└── data/               ← Need read (or read/write) access
    ├── trade/
    └── labour/
```

**Read access** = can query data
**Write access** = can modify data

The permissions work at the file/folder level, just like before.

For PostgreSQL catalogs, you'd also need database credentials - but the Parquet data files still use file system permissions.

---

## Public Catalog: Organisation-wide Discovery

A common challenge: **How do people discover what data exists without having access to all of it?**

In Hive mode, if you can't access the Trade folder, you can't see what datasets Trade has. `datapond` solves this with a **public catalog** - a shared location for metadata only.

### How It Works

```
//CSO-NAS/DataLake/
├── _catalog/             ← Everyone can read this folder
│   ├── Trade/
│   │   ├── Imports.json  ← Metadata only (description, owner, columns)
│   │   └── Exports.json
│   └── Labour/
│       └── Employment.json
├── Trade/                ← Only Trade section can access
│   └── Imports/
│       ├── _metadata.json ← Source metadata
│       └── year=2024/
│           └── data.parquet
└── Labour/               ← Only Labour section can access
    └── Employment/
```

### Publishing to the Public Catalog

```{r}
# Make a dataset discoverable
db_describe(
  section = "Trade",
  dataset = "Imports",
  description = "Monthly import values by country",
  owner = "Trade Section",
  public = TRUE  # Copies metadata to _catalog/Trade/Imports.json
)

# Or use the convenience function
db_set_public(section = "Trade", dataset = "Imports")
```

### Discovering Data

```{r}
# Anyone can see what datasets exist (even without data access)
db_list_public()
#>   section dataset            description        owner
#> 1   Trade Imports Monthly import values  Trade Section
#> 2  Labour Employment    Employment stats Labour Section

# Find datasets by keyword
db_search("monthly", field = "description")
```

### Security Model

| Location | Contains | Who Can Access |
|----------|----------|----------------|
| `_catalog/` | Metadata only (JSON) | Everyone (read) |
| `Trade/` | Actual data (Parquet) | Trade section only |

**The data stays secure** - the public catalog only contains descriptions, not the actual data values.

---

## DuckLake Access Control

DuckLake **automatically organises data** into `{schema}/{table}/` folder structure, enabling folder-level permissions without any extra configuration.

### Architecture

```
//CSO-NAS/DataLake/
├── catalog.sqlite           ← Single DuckLake catalog
└── data/                    ← Data organised automatically by schema/table
    ├── trade/               ← Trade team has access (folder ACLs)
    │   ├── imports/
    │   │   └── ducklake-uuid.parquet
    │   └── exports/
    │       └── ducklake-uuid.parquet
    ├── labour/              ← Labour team has access (folder ACLs)
    │   └── employment/
    │       └── ducklake-uuid.parquet
    └── reference/           ← Everyone has read access
        └── countries/
            └── ducklake-uuid.parquet
```

### Setting Up Schemas

```{r}
# Connect to DuckLake
db_lake_connect(
  catalog_type = "sqlite",
  metadata_path = "//CSO-NAS/DataLake/catalog.sqlite",
  data_path = "//CSO-NAS/DataLake/data"
)

# Create schemas - DuckLake creates folders automatically
db_create_schema("trade")
db_create_schema("labour")
db_create_schema("reference")

# Write data - files go to data/{schema}/{table}/ automatically
db_lake_write(imports_data, schema = "trade", table = "imports")
db_lake_write(countries, schema = "reference", table = "countries")

# Set folder ACLs on the schema folders:
# - //CSO-NAS/DataLake/data/trade/     → Trade team read/write
# - //CSO-NAS/DataLake/data/labour/    → Labour team read/write
# - //CSO-NAS/DataLake/data/reference/ → Everyone read
```

### Using Schemas

```{r}
# Read from a schema (requires folder access)
imports <- db_lake_read(schema = "trade", table = "imports")

# Discover available tables
db_list_schemas()
db_tables("trade")

# Optionally partition for finer access control
db_set_partitioning(schema = "trade", table = "imports", partition_by = c("year"))
db_get_partitioning(schema = "trade", table = "imports")
```

### How It Works

| Component | Access Controlled By |
|-----------|---------------------|
| Catalog file (`.sqlite`) | File permissions - need read to query, write to modify |
| Schema data folder | Folder ACLs - each schema can have different permissions |
| Table data | Inherited from schema folder |

**Benefits**:
- **Zero configuration** - folder structure created automatically
- **Familiar model** - uses standard folder permissions
- **Granular control** - different teams can own different schemas
- **Single catalog** - one metadata file, simpler management
- **IT-friendly** - works with existing permission infrastructure

---

## Putting It Together

Here's how a typical workflow might look:

### Publishing Data (Producer)

```{r}
library(datapond)

# Connect to DuckLake
db_lake_connect(
  catalog_type = "sqlite",
  metadata_path = "//CSO-NAS/DataLake/catalog.sqlite",
  data_path = "//CSO-NAS/DataLake/data"
)

# Prepare your data
imports_q1 <- prepare_imports_data(raw_files)

# Preview what will happen
db_preview_lake_write(imports_q1, schema = "trade", table = "imports", mode = "append")

# Publish with a meaningful commit message
db_lake_write(
  imports_q1,
  schema = "trade",
  table = "imports",
  mode = "append",
  commit_author = Sys.info()["user"],
  commit_message = "Q1 2025 imports data - final"
)

# Document for discovery
db_describe(
  schema = "trade",
  table = "imports",
  description = "Monthly import values by country and HS code",
  owner = "Trade Section"
)

db_disconnect()
```

### Consuming Data (Consumer)

```{r}
library(datapond)

# Connect to DuckLake (read access to catalog and data folder)
db_lake_connect(
  catalog_type = "sqlite",
  metadata_path = "//CSO-NAS/DataLake/catalog.sqlite",
  data_path = "//CSO-NAS/DataLake/data"
)

# Discover what's available
db_list_schemas()
db_tables("trade")

# Search for relevant data
db_search("imports")

# Check documentation
db_get_docs(schema = "trade", table = "imports")

# Read and analyse
imports <- db_lake_read(schema = "trade", table = "imports")

imports |>
  filter(year == 2025, quarter == 1) |>
  summarise(total_value = sum(value)) |>
  collect()

db_disconnect()
```

---

## Glossary

| Term | Meaning |
|------|---------|
| **Parquet** | Columnar file format for storing tabular data efficiently |
| **Hive partitioning** | Organising files into folders named by column values (e.g., `year=2024/`) |
| **Partition pruning** | Query optimisation that skips irrelevant partition folders |
| **DuckLake** | Metadata layer that adds versioning and transactions to Parquet files |
| **Catalog** | Database storing DuckLake metadata (can be DuckDB, SQLite, or PostgreSQL) |
| **Snapshot** | A point-in-time version of the data in DuckLake |
| **Time travel** | Querying data as it existed at a past version or timestamp |
| **Schema** | A namespace for organising related tables; can have custom data paths for access control |
| **Schema path** | Custom data folder for a DuckLake schema, enabling folder-based access control |
| **ACID** | Atomicity, Consistency, Isolation, Durability - database reliability guarantees |
| **Upsert** | Update existing rows + insert new rows in one operation |
| **Data dictionary** | Documentation of all datasets, their columns, types, and descriptions |
| **Public catalog** | Shared location for metadata in hive mode, enabling discovery without data access |
| **Section** | An organisational unit (e.g., Trade, Labour) - folder in hive mode |

---

## Next Steps

- See `vignette("code-walkthrough")` for detailed explanation of how the package code works
- Try the examples in the README to get hands-on experience
- Start with Hive mode if migrating from SAS, move to DuckLake when ready
- For DuckLake, use SQLite catalog on shared drives
